# Volatility Models

```{r setup4, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, warning=FALSE, message=FALSE)
options(digits=4, width=60)
library(fpp3)
library(patchwork)
library(purrr)
library(tidyverse)

```

\pagenumbering{arabic}

## Introduction

- Anything that is observed sequentially over time is a time series.

<!--Tsay-->
- **Financial time series** analysis focuses on the theory and practice of asset valuation over time.
     
<!--Asset valuation is the process of determining the current value of a company's assets, such as stocks, buildings, equipment, brands, goodwill, etc.

https://www.quora.com/What-is-the-difference-between-time-series-methods-and-financial-time-series-methods
As the names suggest, Financial TS methods are a subset of TS methods.

At the end, Financial TS is times series analysis trying to capture different aspects of financial data that we do not see in general in other economics data.

-->
- In finance, the data can be collected much more frequently -- High frequency data.

<!--Hence, we may need to use continuous time econometrics instead of discrete time.-->
- Many financial time series also exhibit changing variance and this can have important consequences
in formulating  financial decisions.

<!--Another famous extension in finance is time varying variance and covariances (GARCH, Stochastic Volatility etc).-->


**Example: Financial time series**

- Typically, when we analyze assets, we look at the percentage change in prices or returns.

```{r sp500, cache=TRUE, echo=TRUE, fig.cap="Daily returns of the adjusted closing prices of the S&P500 index from January 4, 1995 to February 25, 2021"}
# Tidy financial analysis 
library(tidyquant)

sp500 <- tq_get("^GSPC", from = "1995-01-04", to = "2021-02-25" )
print(sp500)

# Convert each assets raw adjusted closing prices to returns
sp500_return <- sp500 %>% 
  tq_transmute(select     = adjusted, 
               mutate_fun = periodReturn, 
               period     = "daily")

sp500_return %>% 
  as_tsibble(index = date) %>%
  autoplot(daily.returns) +
  labs(x = "Day", y= "Daily return")

```






## References:

- Chatfield, C., & Xing, H. (2019). The analysis of time series: an introduction with R. CRC press.

\newpage
