# Volatility Models

```{r setup4, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,  warning=FALSE, message=FALSE, fig.pos='h')
options(digits=4, width=60)
library(fpp3)
library(patchwork)
library(purrr)
library(tidyverse)

```

\pagenumbering{arabic}

**This chapter is heavily based on Chapter 12 of @chatfield2019analysis.**

## Introduction

- Anything that is observed sequentially over time is a time series.

<!--Tsay-->
- **Financial time series** analysis focuses on the theory and practice of asset valuation over time.
     
<!--Asset valuation is the process of determining the current value of a company's assets, such as stocks, buildings, equipment, brands, goodwill, etc.

https://www.quora.com/What-is-the-difference-between-time-series-methods-and-financial-time-series-methods
As the names suggest, Financial TS methods are a subset of TS methods.

At the end, Financial TS is times series analysis trying to capture different aspects of financial data that we do not see in general in other economics data.

-->
- In finance, the data can be collected much more frequently -- High frequency data.
<!-- High frequency data provide a rich source of information on the micro-structure of financial markets.-->
<!--Hence, we may need to use continuous time econometrics instead of discrete time.-->
- Many financial time series also exhibit changing variance and this can have important consequences in formulating  financial decisions.

<!--Another famous extension in finance is time varying variance and covariances (GARCH, Stochastic Volatility etc).-->


**Example: Financial time series**

- Typically, when we analyze assets, we look at the percentage change in prices or returns.

```{r sp500, cache=TRUE, echo=TRUE, fig.cap="Daily returns of the adjusted closing prices of the Standard & Poor's 500 (S&P500) index from January 4, 1995 to February 25, 2021"}
# Tidy financial analysis 
library(tidyquant)

sp500 <- tq_get("^GSPC", from = "1995-01-04", to = "2021-02-25" )
print(sp500)

# Convert each assets raw adjusted closing prices to returns
sp500_return <- sp500 %>% 
  tq_transmute(select     = adjusted, 
               mutate_fun = periodReturn, 
               period     = "daily")

sp500_return %>% 
  as_tsibble(index = date) %>%
  autoplot(daily.returns) +
  labs(x = "Day", y= "Daily return")

```

   - The mean of the return series seems to be stable with an average return of approximately zero.
   - The volatility of data changes over time.

- The focus of this chapter is to study some methods and econometric models for modeling the **volatility** (conditional standard deviation) of an asset return. <!--tsay Volatility means the conditional standard deviation of the underlying asset return.-->
- These models are referred to as **conditional heteroscedastic models**.
<!--https://stats.stackexchange.com/questions/41509/what-is-the-difference-between-garch-and-arma - What is the difference between GARCH and ARMA?-->
- These models do not generally provide better point forecasts, but provides a better estimates of the (local) variance.
- As a result they allow to compute   more reliable prediction intervals and therefore a better assessment of risk. 
- Volatility models have many applications in economics and finance.
- This chapter discusses various types of **univariate volatility models**


## Structure of a Model for Asset Returns

- Let, $\{P_T\}$, denotes a time series.

- Let, $\{Y_T\}$, denotes a derived series from which any trend and seasonal effects have been removed and  linear (short-term correlations) effects may also have been removed.
  - Examples : Let  $\{P_T\}$ be share price at the $t$th trading day.
  
  $$Y_t=log P_t - log P_{t-1} \text{  or  } Y_t = \frac{P_t-P_{t-1}}{P_{t-1} }\times100\%$$
   
<!--$\{Y_T\}$ might be the first differences (or percentage changes) of financial time series such as the natural log of a share price $\{P_T\}$
-->
   - This is often called the **return** or the **growth rate** of a series.

**Example**

- Let $P_t$ be the adjusted closing prices of the S&P500 at the $t$th trading day.
- Let $Y_t$ be the dialy returns of the S&P500 Index at each day as shown in Figure \@ref(fig:sp500).
- The basic idea in volatility modelling is that the return series $\{Y_t\}$ has very few serial correlations, but it is a dependent series.
- Consider the sample ACFs and PACFs of $Y_t$, $|Y_t|$ and $Y_t^2$ (Figure \@ref(fig:acf))


```{r acf, message=FALSE, warning=FALSE, fig.pos='h', fig.cap="Sample ACF (left) and sample PACF (right) of various functions of the daily returns, $Y_t$, of adjusted closing prices of S&P500 Index from from January 4, 1995 to February 25, 2021. Top: Original series $Y_t$; Middle: Absolute value of $Y_t$; Bottom: Squared values of $Y_t$." }
data <- sp500_return %>% 
  mutate(yt= daily.returns, 
         absyt = abs(daily.returns),
         sqyt = daily.returns^2) %>%
  as_tsibble(index = date, regular=FALSE)

p1<- data %>% ACF(yt) %>% autoplot()
p2<- data %>% PACF(yt) %>% autoplot()
p3<- data %>% ACF(absyt) %>% autoplot()
p4<- data %>% PACF(absyt) %>% autoplot()
p5<- data %>% ACF(sqyt) %>% autoplot()
p6<- data %>% PACF(sqyt) %>% autoplot()

(p1|p2)/
  (p3|p4)/
  (p5|p6)
```

  - Sample ACFs the returns $Y_t$ suggest no significant serial correlations except for small ones at lags 1, 3 and 5.
  - However, the sample ACFs of $|Y_t|$ and  $Y_t^2$, show strong dependence over all lags.
  - Important feature: the returns may seem serially uncorrelated, but it is dependent.
  - This is a common observations for daily returns series
  
<!-- 
https://stats.stackexchange.com/questions/325390/serially-uncorrelated-but-dependence-in-arch-model

Serially Uncorrelated but dependence in ARCH model

On the first chart (of raw returns), the ACF does not appear to be significantly non-0 anywhere (other than at lag 1 of course). In other words, there is no serial correlation of the returns.

On the second chart (of squared returns), the ACF does appear to be significantly non-0 at certain lags. So this gives us hope that we can use these squared returns to predict something by using them.

(Interestingly, if you plotted the ACF for absolute returns, you would find something similar to the second chart. This is because the absolute value and the square both discard the sign to measure some sort of "deviation", in the non-technical sense of the word.)

It's not just any series, it's a series of raw returns. And as you saw in the first chart, raw returns are not serially correlated.

However, squared (and absolute) returns are, which is good news! This is because we can now use them to predict "volatility" (i.e. the conditional variance) using ARCH models. And this is what ARCH models do-->


<!-- To focus our discussion on volatility of return series, we may further assume that Yt-->

**Volatility of a return series**

- Let $Y_t$ be the innovations in a linear time series model.
<!--  the innovation is the difference between the observed value of a variable at time t and the optimal forecast of that value based on information available prior to time t.

 I conclude that innovations are ok to interchange with errors. It's called innovations because in time series context the errors bring new information to the system.-->
- Let $X_t$ follow an $ARMA(p,q)$ model,
$$\phi(B)X_t=\theta(B)Y_t,$$

where $\phi(B)$ and $\theta(B)$ are polynomials of $B$ with order $p$ and $q$, respectively.
- Let $\mathcal{F}_t$, the set of observed data upto time $t$, (i.e. $\{X_1,X_2,\dots, X_t\}$).
- Then the observation $X_t$ can be written as 
$$X_t=\mu_t+Y_t,$$
where $\mu_t$ is the mean of $X_t$ conditional on observed data $\mathcal{F}_{t-1},$

$$\mu_t=E(X_t|\mathcal{F}_{t-1})= \phi(B)X_t-(\theta(B)-1)Y_t$$

## References:

- Chatfield, C., & Xing, H. (2019). The analysis of time series: an introduction with R. CRC press.

- Tsay, R. S. (2005). Analysis of financial time series (Vol. 543). John wiley & sons.
\newpage
